Measuring contrast processing in the visual system using the steady state visually evoked potential (SSVEP)

### 2025-01-30 review.tex
Getting there... replaced  .pdf with .png but still not quite working

Neurons in the visual areas of the brain are primarily responsive to
changes in cone photoreceptor activations across time and space. This
property, referred to as `contrast', sets the fundamental limits of our
visual abilities, which remain steady over a remarkably wide range of
environmental light levels. The human response to contrast can be
studied using many different techniques. Early work used psychophysical
methods to measure contrast sensitivity Schade (1956), defined as the
inverse of the lowest contrast that can be reliably detected. But neural
responses can also be measured more directly using techniques such as
functional magnetic resonance imaging (fMRI), magnetoencephalography
(MEG), and electroencephalography (EEG). Here we will describe how an
EEG method known as the steady state visually evoked potential (SSVEP)
technique has contributed to our understanding of human contrast
processing in health, disease and throughout development.

@@

Two common stimulus variants involve sinusoidal on-off (or
`appearance/disappearance') flicker, where the stimulus alternates
between a blank background and the peak contrast, and sinusoidal
counterphase flicker, where the stimulus alternates in phase (i.e the
black regions become white and the white regions become black). On-off
flicker can excite independent populations of on- and off-cells in the
visual system once per cycle. For spatial patterns, the contributions of
individual cortical cells to these excitations are thought to sum at the
scalp and generate an average of their onset responses on each cycle.
For very low spatial frequencies (including zero) resposes at the scalp
will be dominated by a single on- or off- cell type on each half cycle.

@@

By comparison, counterphase flickering patterns generate two essentially
identical transients per cycle and therefore does not produce a response
at 1F, only even harmonics: 2F, 4F, 6F and so on. Because square-waves
are spectrally broad-band, square wave flicker (either on/off or
contrast reversig) tends to produce additional spectral harmonics
compared to sine-wave flicker.

The higher harmonics of the steady-state signal are generally thought to
reflect nonlinear processing in the visual system (Regan \& Regan,
1988). Because they can arise from different neuronal computations,
different populations and even different stages of the visual system,
different harmonics can exhibit different input-output functions. For
example, 1F and 2F responses can have different thresholds (Bobak
\emph{et al.}, 1984) and scalp topographies (Regan, 1973) while Kaester
et al (Kaestner \emph{et al.}, 2024) show both different response slopes
and thresholds for 1F and 2F components generated by dynamic noise.

@@

Contrast is one of the most fundamental pieces of information that the
eye transmits to the brain. It can be defined as the change in cone
photoreceptor activity over space (`spatial contrast') or time
(`temporal contrast'). Cone photoreceptors - which drive precortical
opponent pathways - contribute to both chromatic and achromatic
contrast, and although most of the research we describe here focuses on
achromatic contrast, SSVEPs have proven to be an excellent measure of
early chromatic processing as well McKeefry \emph{et al.} (1996) (see
also (Baseler \& Sutter, 1997)).

@@

Although its mathematical definition is straightforward, the
computations that underlie contrast processing in the brain have been
the subject of intense research for many decades. The neural code for
contrast, even in the earliest parts of visual cortex, is not simply a
linear transform of the contrast at the retina - instead, contrast
signals undergo a cascade of nonlinear processing stages that, broadly,
attempt to normalise the output relative to the spatiotemporal
environment. This normalization, achieved through a computation called
`contrast gain control' (Heeger, 1992; Foley, 1994; Carandini \& Heeger,
2011) maximises the sensitivity of the visual system by making optimal
use of neuronal bandwidth. As an example, a grating placed at the centre
of a low-contrast background typically appears more intense than the
same grating when superimposed on a high contrast background (see
Figure~\ref{fig-centresurround}; note that the code used to produce all
figures in this review is available in python and R at:
https://github.com/wadelab/contrastReviewPaperVNS).

@@

\includegraphics{Figures/1_centresurround.pdf}

@@

Although the SSVEP does seem to reflect the activity of relatively
well-tuned neuronal populations , changing some aspect of the stimulus
may change the nature of those populations. For example, Campbell and
Maffei (Campbell \& Maffei, 1970) noted that densely-sampled
measurements of contrast responses can reveal the presence of two,
qualitatively different types of neurons that exhibit different
log-linear contrast functions (see also Souza \emph{et al.} (2007)).
Modeling such two-limbed contrast response functions using a single
sigmoidal function is therefore an approximation.

@@

An additional complexity is introduced by the fact that the EEG is an
average population response and that the visual system contains many
different types of neurons. As a stimulus changes in, for example,
contrast it may selectively activate qualitatively different neuronal
populations. This type

To understand these processes better, we will show how sinusoidal input
signals might be processed by the visual system to produce SSVEPs.
Figure~\ref{fig-linear} illustrates how sine waves of different
contrasts are processed in a linear system. The first panel shows the
input sine wave, which would be used to modulate stimulus amplitude over
time. Notice that there are five peaks in the waveform during the one
second sample, so the stimulation frequency is 5Hz (F1). The second
panel shows Fourier transform of the waveform, which contains a
substantial peak at this frequency. If we change the stimulus contrast
(i.e.~the amplitude of the waveform), the amplitude of the F1 component
increases linearly with contrast (right panel).

@@

First, we consider the fact that neurons in the early visual system code
either positive or negative contrast. The responses of each different
type of neuron is an increase in firing rate to the increasing levels of
the prefered contrast polatiry and these neurons typically exhibit zero
(or baseline) response to the non-prefered polarity. We can model the
combined responses of these two cell populations to a time varying
contrast signal by full-wave-rectification. The effect of this
computation is, effectively, to double the input freuquency in the
population measurement and to introduce additional higher (even)
harmonics due to the discontinuity at the contrast reversals. The effect
of full-wave rectification on the SSVEP signal is shown in
Figure~\ref{fig-fullwaverectification}

@@

One of the simplest nonlinearities is the function that describes a
cell's response to different levels of contrast. In
Figure~\ref{fig-transducer} this is modeled by a hyperbolic ratio
function resulting in a saturating non-linearity:

@@

\includegraphics{Figures/4_transducer.pdf_transducer.pdf}

@@

To illustrate the effect of contrast gain control (Heeger, 1992), we
next include a second component (a `mask') that contributes to the gain
pool of the first (the `target'). The mask stimulus will suppress the
response at the target frequency, reducing its amplitude. The
suppression is reciprocal - activity at the mask frequency is also
reduced by the presence of the target, in a contrast-dependent manner
Busse \emph{et al.} (2009). At a single pair of (matched) contrast
levels, we see a complex pattern of intermodulation terms in the
transduced waveforms (left panel of Figure~\ref{fig-twoinputs}) and in
the Fourier spectrum (middle panel of Figure~\ref{fig-twoinputs}). The
two components interact generating nonlinear intermodulation terms at
sums and differences of the input frequencies (F1, F2). The nature of
the interaction --- and therefore the pattern of intermodulation terms
--- is determined by the computations happening as the contrast signal
moves from retina to cortex (Regan \& Regan, 1988).

@@

\includegraphics{Figures/4_twoinputs.pdf}

@@

where \(C_\mathrm{mask}\) reflects the contrast of the mask component at
a distinct frequency from that of the target. The effect of this extra
term is to reduce the target response (see right panel of
Figure~\ref{fig-twoinputs}). For a linear contrast axis, the contrast
response function becomes shallower, whereas on a logarithmic contrast
axis it maintains its steepness and shifts to the right. Suppressive
effects of this kind have been obtained using SSVEP with a variety of
different types of mask, including orthogonal overlaid masks, surround
masks, and dichoptic masks (Burr \& Morrone, 1987; Ross \& Speed, 1991;
Candy \emph{et al.}, 2001; Busse \emph{et al.}, 2009; Cunningham
\emph{et al.}, 2017; Salelkar \& Ray, 2020; for a meta-analysis see
Baker \emph{et al.}, 2021).

Even without considering a spatial component, the early visual system is
far more complex than the model here suggests. For example as well as
cells that code positive or negative contrast in a more or less
continuous manner, the retina also contains `transient' cells that code
temporal changes in contrast. These cells (Kuffler \emph{et al.}, 1957;
Alpern, 1971), and cells with similar properties in the lateral
geniculate nucleus (LGN) (Levitt \emph{et al.}, 2001) and cortex (Hubel
\& Wiesel, 1959; Movshon, 1975), will introduce second harmonic
components even when the stimulus itself is modulated in an on-off
fashion (McKeefry \emph{et al.}, 1996). Analogously, in the spatial
domain, so-called `simple cells' are sensitive to the polarity of a
spatial contrast modulation while `complex cells' respond to the
presence of patterned spatial contrast irrespective of its spatial
position (Hubel \& Wiesel, 1962). Both the amplitude and the phase of
the SSVEP response to a single contrast-reversing sine-wave grating
therefore contains information about nonlinear computations performed
across a range of retinal and cortical cell types.

The complexity of even a simple simulation of the frequency-domain
signal arising from non-linear interactions is intriguing. Presumably,
the signal measured from early visual cortex is the result of a cascade
of nonlinear retinal and cortical operations up to that point. It
therefore contains a `signature' or `fingerprint' of the computational
nature, order and parameters of those operations - including the shape
of the transducer functions and the computations involved in signal
combination. In principle, that information could be recovered from the
SSVEP signal - a possibility recognised in the early days of the
technique (Regan \& Regan, 1988). Although characterising the complete
set of computations along entire processing pathway is challenging, and
a full system identification may require the use of multiple different
input frequencies (Boyd \emph{et al.}, 1983), careful parametric
variation of the input stimuli does allow us to fit models of early
visual visual processing (e.g. (Zemon \& Gordon, 2006)) and, by
incorporating mathematical models of gain control mechanisms with either
instantaneous, or temporally-integrated gain control, it is possible to
model gain control using SSVEP data. For example, (Tsai \emph{et al.},
2012) demonstrated that a relatively simple gain control model gave a
good account of the pattern of intermodulation responses produced by two
overlaid patterns flickering at different frequencies. This was achieved
by passing full stimulus waveforms through the transducer nonlinearity,
and calculating the Fourier spectrum of the model output. More recent
work has used this approach to study binocular integration in both
normal subjects and amblyopesHou \emph{et al.} (2020). Our more recent
work on signal combination across eyes and space similarly demonstrated
close correspondence between the predictions of a computational model
and empirical data in humans (Baker \& Wade, 2017). More detailed
modelling of intracortical recordings (Groen \emph{et al.}, 2022) has
revealed details of the timecourse of gain control effects, specifically
that normalization is delayed slightly relative to the initial visual
response.

Similar changes to the contrast response function might also be obtained
using adaptation paradigms, in which the visual system is exposed to
high contrast stimuli for long durations. Psychophysically, adaptation
increases detection thresholds, but has little effect on contrast
discrimination performance (Ross \emph{et al.}, 1993), much like pattern
masks (Foley, 1994). Although SSVEP adaptation effects show strong
tuning for orientation (Campbell \& Maffei, 1970; Vergeer \emph{et al.},
2018) and spatial frequency (Mecacci \& Spinelli, 1976), there appear to
be very measurements of the full SSVEP contrast response function before
and after an extended period of adaptation. One exception is a study by
Bach et al (Bach \emph{et al.}, 1988) that examined the effects of
adaptation on SSVEP responses who found a paradoxical increase in the 1F
SSVEP amplitude following prolonged exposure to high-contrast stimuli
across much of the probe contrast range. They explain this as being due
to the presence of a nonlinear transducer function that saturates at
high probe contrast driving the response to higher harmonic frequencies.
Adaptation reduces the apparent contrast of the input, leading to less
distortion, a more sinusoidal output and a concentration of the
frequency-tagged response at lower harmonics. This explanation is
conceptually related to the `supersaturation' effect that we observe in
Figure 4.

@@

Because SSVEP responses at detection threshold are very small,
estimating a threshold is achieved by measuring the contrast response
function at relatively high levels (in the linear part of the
log-contrast response function prior to saturation), and extrapolating
back along the function (either contrast vs response measured at a
constant spatial frequency or spatial frequency vs response at a
constant contrast level) to estimate its intercept with the x-axis
(Campbell \& Maffei, 1970) (see Figure~\ref{fig-sweepvep}). This
contrast level has been shown to correspond approximately with
psychophysically measured detection thresholds Norcia \emph{et al.}
(1986). One factor affecting the accuracy of this technique is the
presence of noise in the EEG signal which will increase the apparent
amplitude of the SSVEP responses across the measurement range. The
effect of this noise is particulary apparent for recordings with low SNR
or where the slope of the CRF is shallow (and therefore where small
changes in the vertical offset of the fit result in large changes in the
X-axis intercept). Accurate estimates of thresholds therefore require
the experimenter to model this noise and compensate for its effects - as
detailed in (Norcia \emph{et al.}, 1989).

@@

\includegraphics{Figures/5_sweepplot.pdf}

@@

A robust estimate of the threshold therefore requires the measurement of
the SSVEP amplitude at many different super-threshold contrast levels.
This was made faster by the development of the `sweep VEP' paradigm in
which the stimulus changed its contrast, spatial frequency or some other
property, throughout a trial Tyler \emph{et al.} (1979). To avoid
hysteresis or short-term adaptation effects, the sweep is sometimes
conducted both up and down in the same experiment (Tyler \emph{et al.},
1979; Norcia \& Tyler, 1985\emph{a}; Norcia \emph{et al.}, 1990). The
sweep VEP (really, a sweep `SSVEP') technique is now commonly-used to
obtain a rapid and objective measurement of visual acuity. In
particular, because of its relative speed and simplicity, this technique
has now become a standard for conducting tests of visual acuity in very
young subjects or where behavioural tests are not appropriate (Ridder,
2004; Bach \emph{et al.}, 2008; Hoffmann \emph{et al.}, 2017; Bach \&
Farmer, 2020).

The approach of estimating visual thresholds by extrapolating SSVEP
responses at higher stimulus levels has revealed much about the
development of visual abilities in infants (Harris \emph{et al.}, 1976;
Atkinson \emph{et al.}, 1979; Braddick \emph{et al.}, 1986). In general,
SSVEP measurements of infant vision have revealed that contrast
sensitivity and acuity for both achromatic and chromatic stimuli as well
as stereoscopic depth perception develops earlier than had been supposed
previously based on behavioural readouts (Dobson \emph{et al.}, 1978;
Norcia \& Tyler, 1985\emph{b}). For example, using SSVEPs, Norcia and
Tyler (Norcia \& Tyler, 1985\emph{b}) found acuity thresholds in one
month old infants that were roughly half those estimated previously from
behavioural experiments (reviewed in (Brown, 1990)). The ability to make
very sensitive measurements of early visual function in infants also
allows researchers to monitor the trajectory of visual development and
identify different temporal epochs of developmental change (Norcia
\emph{et al.}, 1990).

SSVEP measures of thresholds for detecting purely chromatic stimuli find
responses to isoluminant stimuli by around 5 to six weeks with contrast
detection reaching near-adult levels by around six months (Morrone
\emph{et al.}, 1993). Spatial acuity as measured by SSVEP reaches adult
levels more slowly, but near-adult levels are recorded around one year
(Norcia \& Tyler, 1985\emph{b}; Hamer \emph{et al.}, 1989) compared to
around six to seven years with behavioural measures (Atkinson \&
Braddick, 1983; Ellemberg \emph{et al.}, 1999).

At least some of this difference is likely due to the relative
objectivity and high SNR of the SSVEP technique compared to other
methods such as preferential looking, which require careful measurement
of the infant's gaze direction. In slightly older chilren, other groups
have reported electrophysiological correlates of visual acuity that more
closely match the behavioural measures (De Vries-Khoe \& Spekreijse,
1982).

The SSVEP technique has also been used to study the development of the
contrast gain control mechanisms described in the previous section.
Although response- (Morrone \& Burr, 1986) or contrast gain control
{[}Candy \emph{et al.} (2001); ; Skoczenski \& Norcia (1998){]} is
measurable in infants as young as six weeks old, its development appears
to be slower, with adult levels being reached at approximately 11 years
(Pei \emph{et al.}, 2017).

@@

Amblyopia is a disorder of binocular vision, characterised by one eye
contributing much less to perception than the other. This is often due
to strabismus (squint) or anisometropia (difference in optical
prescription between the eyes) during development. Contemporary accounts
suggest that the amblyopic eye is suppressed by signals from the fellow
eye. SSVEPs provide a convenient and objective method to characterise
the difference in neural response to signals in each eye, and typically
show reduced responses to stimuli in the amblyopic eye (see
Figure~\ref{fig-plotclinical}b) across the contrast range (Baker
\emph{et al.}, 2015; Lygo \emph{et al.}, 2021). Measurments of
interocular contrast modulation (for example, Hou 侯川 \emph{et al.}
(2021) can also provide information about the way that amblyopia changes
the balance of excitation and inhibition between eyes.

There are currently many novel binocular treatments for amblyopia under
development, often involving virtual reality or stereo display systems
designed to encourage the two eyes to work together. The steady-state
approach may be more sensitive and objective than typical acuity
measurements, and also has the potential to measure suppression between
the eyes directly (e.g. Zheng \emph{et al.}, 2019; Hu \emph{et al.},
2023; Du \emph{et al.}, 2023).

@@

\includegraphics{Figures/6_clinicalplot.pdf}

@@

@@

@@

\bibitem[\citeproctext]{ref-reganRapidObjectiveRefraction1973}
REGAN D (1973). Rapid {Objective Refraction Using Evoked Brain
Potentials}. \emph{Investigative Ophthalmology \& Visual Science}
\textbf{12,} 669--679.

>>>>

Neurons in the visual areas of the brain are primarily responsive to
changes in cone photoreceptor activations across time and space. This
property, referred to as `contrast', sets the fundamental limits of our
visual abilities, which remain steady over a remarkably wide range of
environmental light levels. The human response to contrast can be
studied using many different techniques. Early work used psychophysical
methods to measure contrast sensitivity (Schade, 1956; Campbell \&
Green, 1965), defined as the inverse of the lowest contrast that can be
reliably detected. But neural responses can also be measured more
directly using techniques such as functional magnetic resonance imaging
(fMRI), magnetoencephalography (MEG), and electroencephalography (EEG).
Here we will describe how an EEG method known as the steady state
visually evoked potential (SSVEP) technique has contributed to our
understanding of human contrast processing in health, disease and
throughout development.

@@

Two common stimulus variants involve sinusoidal on-off (or
`appearance/disappearance') flicker, where the stimulus alternates
between a blank background and the peak contrast, and sinusoidal
counterphase flicker, where the stimulus alternates in phase (i.e the
black regions become white and the white regions become black). On-off
flicker can excite independent populations of on- and off-cells in the
visual system once per cycle. For spatial patterns, the contributions of
individual cortical cells to these excitations are thought to sum at the
scalp and generate an average of their onset responses on each cycle.
For very low spatial frequencies (including zero), resposes at the scalp
will be dominated by a single on- or off- cell type on each half cycle.

@@

By comparison, counterphase flickering patterns generate two essentially
identical transients per cycle and therefore do not produce a response
at 1F, only even harmonics: 2F, 4F, 6F and so on. Because square-waves
are spectrally broad-band, square wave flicker (either on/off or
contrast reversing) tends to produce additional spectral harmonics
compared to sine-wave flicker.

The higher harmonics of the steady-state signal are generally thought to
reflect nonlinear processing in the visual system (Regan \& Regan,
1988). Because they can arise from different neuronal computations,
different populations and even different stages of the visual system,
different harmonics can exhibit different input-output functions. For
example, 1F and 2F responses can have different thresholds (Bobak
\emph{et al.}, 1984) and scalp topographies (Regan, 1973\emph{a}) while
Kaestner \emph{et al.} (2024) show both different response slopes and
thresholds for 1F and 2F components generated by dynamic noise.

@@

Contrast is one of the most fundamental pieces of information that the
eye transmits to the brain. It can be defined as the change in cone
photoreceptor activity over space (`spatial contrast') or time
(`temporal contrast'). Cone photoreceptors - which drive precortical
opponent pathways - contribute to both chromatic and achromatic
contrast, and although most of the research we describe here focuses on
achromatic contrast, SSVEPs have proven to be an excellent measure of
early chromatic processing as well (Regan, 1966\emph{b}, 1973\emph{a},
1975; McKeefry \emph{et al.}, 1996; Di Russo \emph{et al.},
2001\emph{a}) (see also Baseler \& Sutter, 1997).

@@

Although its mathematical definition is straightforward, the
computations that underlie contrast processing in the brain have been
the subject of intense research for many decades. The neural code for
contrast, even in the earliest parts of visual cortex, is not simply a
linear transform of the contrast at the retina - instead, contrast
signals undergo a cascade of nonlinear processing stages that, broadly,
attempt to normalise the output relative to the spatiotemporal
environment. This normalization, achieved through a computation called
`contrast gain control' (Ohzawa \emph{et al.}, 1982, 1985; Bobak
\emph{et al.}, 1988; Heeger, 1992; Foley, 1994; Carandini \& Heeger,
2011) maximises the sensitivity of the visual system by making optimal
use of neuronal bandwidth. As an example, a grating placed at the centre
of a low-contrast background typically appears more intense than the
same grating when superimposed on a high contrast background (see
Figure~\ref{fig-centresurround}; note that the code used to produce all
figures in this review is available in python and R at:
https://github.com/wadelab/contrastReviewPaperVNS).

@@

\includegraphics{Figures/1_centresurround.png}

@@

Although the SSVEP does seem to reflect the activity of relatively
well-tuned neuronal populations, changing some aspect of the stimulus
may change the nature of those populations. For example, Campbell \&
Maffei (1970) noted that densely-sampled measurements of contrast
responses can reveal the presence of two qualitatively different types
of neurons that exhibit different log-linear contrast functions
(Nakayama \& Mackeben, 1982; see also Tyler \& Apkarian, 1985; Norcia
\emph{et al.}, 1989; Souza \emph{et al.}, 2007). Modeling such
two-limbed contrast response functions using a single sigmoidal function
is therefore an approximation.

@@

An additional complexity is introduced by the fact that the EEG is an
average population response and that the visual system contains many
different types of neurons. As a stimulus changes in (for example)
contrast, it may selectively activate qualitatively different neuronal
populations.

To understand these processes better, we will show how sinusoidal input
signals might be processed by the visual system to produce SSVEPs.
Figure~\ref{fig-linear} illustrates how sine waves of different
contrasts are processed in a linear system. The first panel shows the
input sine wave, which would be used to modulate stimulus amplitude over
time. Notice that there are five peaks in the waveform during the one
second sample, so the stimulation frequency is 5Hz (F1). The second
panel shows the Fourier transform of the waveform, which contains a
substantial peak at this frequency. If we change the stimulus contrast
(i.e.~the amplitude of the waveform), the amplitude of the F1 component
increases linearly with contrast (right panel).

@@

First, we consider the fact that neurons in the early visual system code
either positive or negative contrast. The response of each different
type of neuron is an increase in firing rate to the increasing levels of
the preferred contrast polarity, and these neurons typically exhibit
zero (or baseline) response to the non-preferred polarity. We can model
the combined responses of these two cell populations to a time varying
contrast signal by full-wave-rectification. The effect of this
computation is, effectively, to double the input frequency in the
population measurement and to introduce additional higher (even)
harmonics due to the discontinuity at the contrast reversals. The effect
of full-wave rectification on the SSVEP signal is shown in
Figure~\ref{fig-fullwaverectification}

@@

The contrast response function (CRF) is still linear for this model -
rectification has doubled the frequency (and introduced higher
harmonics) but the response amplitude measured by the second harmonic is
still directly proportional to the input amplitude.

Neurons in the visual system typically do not exhibit this perfect
response linearity. Instead, neuronal response are well-modeled by some
form of saturating non-linearity: as the contrast increases, the
response of the neuron increases but the response increase per unit
contrast is reduced at the high end of the contrast range. This
saturating nonlinearity is often modeled as a hyperbolic ratio function
of the form:

@@

\includegraphics{Figures/4_transducer.pdf}

@@

To illustrate the effect of contrast gain control (Ohzawa \emph{et al.},
1982; Heeger, 1992), we next include a second component (a `mask') that
contributes to the gain pool of the first (the `target'). The mask
stimulus will suppress the response at the target frequency, reducing
its amplitude. The suppression is reciprocal - activity at the mask
frequency is also reduced by the presence of the target, in a
contrast-dependent manner (Regan \& Regan, 1988; Brown \emph{et al.},
1999; Candy \emph{et al.}, 2001; Busse \emph{et al.}, 2009). At a single
pair of (matched) contrast levels, we see a complex pattern of
intermodulation terms in the transduced waveforms (left panel of
Figure~\ref{fig-twoinputs}) and in the Fourier spectrum (middle panel of
Figure~\ref{fig-twoinputs}). The two components interact generating
nonlinear intermodulation terms at sums and differences of the input
frequencies (F1, F2). The nature of the interaction --- and therefore
the pattern of intermodulation terms --- is determined by the
computations happening as the contrast signal moves from retina to
cortex (Regan \& Regan, 1988).

@@

\includegraphics{Figures/5_twoinputs.pdf}

@@

where \(C_\mathrm{mask}\) reflects the contrast of the mask component at
a frequency distinct from that of the target. The effect of this extra
term is to reduce the target response (see right panel of
Figure~\ref{fig-twoinputs}). For a linear contrast axis, the contrast
response function becomes shallower, whereas on a logarithmic contrast
axis it maintains its steepness and shifts to the right. Suppressive
effects of this kind have been obtained using SSVEP with a variety of
different types of mask, including orthogonal overlaid masks, surround
masks, and dichoptic masks (Burr \& Morrone, 1987; Ross \& Speed, 1991;
Candy \emph{et al.}, 2001; Busse \emph{et al.}, 2009; Cunningham
\emph{et al.}, 2017; Salelkar \& Ray, 2020; for a meta-analysis see
Baker \emph{et al.}, 2021).

Even without considering a spatial component, the early visual system is
far more complex than the model here suggests. For example as well as
cells that code positive or negative contrast in a more or less
continuous manner, the retina also contains `transient' cells that code
temporal changes in contrast. These cells (Kuffler \emph{et al.}, 1957;
Alpern, 1971), and cells with similar properties in the lateral
geniculate nucleus (LGN) (Levitt \emph{et al.}, 2001) and cortex (Hubel
\& Wiesel, 1959; Movshon, 1975), will introduce second harmonic
components even when the stimulus itself is modulated in an on-off
fashion (McKeefry \emph{et al.}, 1996). Analogously, in the spatial
domain, so-called `simple cells' are sensitive to the polarity of a
spatial contrast modulation while `complex cells' respond to the
presence of patterned spatial contrast irrespective of its spatial
position (Hubel \& Wiesel, 1962). Both the amplitude and the phase of
the SSVEP response to a single contrast-reversing sine-wave grating
therefore contain information about nonlinear computations performed
across a range of retinal and cortical cell types.

The complexity of even a simple simulation of the frequency-domain
signal arising from non-linear interactions is intriguing. Presumably,
the signal measured from early visual cortex is the result of a cascade
of nonlinear retinal and cortical operations up to that point. It
therefore contains a `signature' or `fingerprint' of the computational
nature, order and parameters of those operations - including the shape
of the transducer functions and the computations involved in signal
combination. In principle, that information could be recovered from the
SSVEP signal - a possibility recognised in the early days of the
technique (Regan \& Regan, 1988). Although characterising the complete
set of computations along the entire processing pathway is challenging,
and a full system identification may require the use of multiple
different input frequencies (Boyd \emph{et al.}, 1983), careful
parametric variation of the input stimuli does allow us to fit models of
early visual visual processing (e.g. Zemon \& Gordon, 2006) and, by
incorporating mathematical models of gain control mechanisms with either
instantaneous, or temporally-integrated gain control, it is possible to
model gain control using SSVEP data. For example, Tsai \emph{et al.}
(2012) demonstrated that a relatively simple gain control model gave a
good account of the pattern of intermodulation responses produced by two
overlaid patterns flickering at different frequencies. This was achieved
by passing full stimulus waveforms through the transducer nonlinearity,
and calculating the Fourier spectrum of the model output. More recent
work has used this approach to study binocular integration in both
normal subjects and amblyopes (Hou \emph{et al.}, 2020; Hou 侯川
\emph{et al.}, 2021). Our more recent work on signal combination across
eyes and space similarly demonstrated close correspondence between the
predictions of a computational model and empirical data in humans (Baker
\& Wade, 2017). More detailed modelling of intracortical recordings
(Groen \emph{et al.}, 2022) has revealed details of the timecourse of
gain control effects, specifically that normalization is delayed
slightly relative to the initial visual response.

Similar changes to the contrast response function might also be obtained
using adaptation paradigms, in which the visual system is exposed to
high contrast stimuli for long durations. Psychophysically, adaptation
increases detection thresholds, but has little effect on contrast
discrimination performance (Ross \emph{et al.}, 1993), much like pattern
masks (Foley, 1994). Although SSVEP adaptation effects show strong
tuning for orientation (Campbell \& Maffei, 1970; Vergeer \emph{et al.},
2018) and spatial frequency (Mecacci \& Spinelli, 1976), there appear to
be very few measurements of the full SSVEP contrast response function
before and after an extended period of adaptation. One exception is a
study by Bach \emph{et al.} (1988) that examined the effects of
adaptation on SSVEP responses, and found a paradoxical increase in the
1F SSVEP amplitude following prolonged exposure to high-contrast stimuli
across much of the probe contrast range. They explain this as being due
to the presence of a nonlinear transducer function that saturates at
high probe contrasts, driving the response to higher harmonic
frequencies. Adaptation reduces the apparent contrast of the input,
leading to less distortion, a more sinusoidal output and a concentration
of the frequency-tagged response at lower harmonics. This explanation is
conceptually related to the `supersaturation' effect that we observed in
Figure 4.

@@

Because SSVEP responses at detection threshold are very small,
estimating a threshold is achieved by measuring the contrast response
function at relatively high levels (in the linear part of the
log-contrast response function prior to saturation), and extrapolating
back along the function (either contrast vs response measured at a
constant spatial frequency or spatial frequency vs response at a
constant contrast level) to estimate its intercept with the x-axis
(Campbell \& Maffei, 1970) (see Figure~\ref{fig-sweepvep}). This
contrast level has been shown to correspond approximately with
psychophysically measured detection thresholds (Campbell \& Kulikowski,
1972; Norcia \emph{et al.}, 1986). One factor affecting the accuracy of
this technique is the presence of noise in the EEG signal which will
increase the apparent amplitude of the SSVEP responses across the
measurement range. The effect of this noise is particulary apparent for
recordings with low SNR or where the slope of the CRF is shallow (and
therefore where small changes in the vertical offset of the fit result
in large changes in the X-axis intercept). Accurate estimates of
thresholds therefore require the experimenter to model this noise and
compensate for its effects, as detailed in Norcia \emph{et al.} (1989).

@@

\includegraphics{Figures/6_sweepplot.pdf}

@@

A robust estimate of the threshold therefore requires the measurement of
the SSVEP amplitude at many different super-threshold contrast levels.
This was made faster by the development of the `sweep VEP' paradigm in
which the stimulus changed its contrast, spatial frequency or some other
property, throughout a trial (Regan, 1973\emph{b}; Tyler \emph{et al.},
1979). To avoid hysteresis or short-term adaptation effects, the sweep
is sometimes conducted both up and down in the same experiment (Tyler
\emph{et al.}, 1979; Norcia \& Tyler, 1985\emph{a}; Norcia \emph{et
al.}, 1990). The sweep VEP (really, a `sweep SSVEP') technique is now
commonly-used to obtain a rapid and objective measurement of visual
acuity. In particular, because of its relative speed and simplicity,
this technique has now become a standard for conducting tests of visual
acuity in very young subjects or where behavioural tests are not
appropriate (Ridder, 2004; Bach \emph{et al.}, 2008; Hoffmann \emph{et
al.}, 2017; Bach \& Farmer, 2020).

The approach of estimating visual thresholds by extrapolating SSVEP
responses at higher stimulus levels has revealed much about the
development of visual abilities in infants (Harris \emph{et al.}, 1976;
Atkinson \emph{et al.}, 1979; Braddick \emph{et al.}, 1986). In general,
SSVEP measurements of infant vision have revealed that contrast
sensitivity and acuity for both achromatic and chromatic stimuli
develops earlier than had been supposed previously based on behavioural
readouts (Dobson \emph{et al.}, 1978; Norcia \& Tyler, 1985\emph{b}).
For example, using SSVEPs, Norcia \& Tyler (1985\emph{b}) found acuity
thresholds in one month old infants that were roughly half those
estimated previously from behavioural experiments (reviewed in Brown,
1990). The ability to make very sensitive measurements of early visual
function in infants also allows researchers to monitor the trajectory of
visual development and identify different temporal epochs of
developmental change (Norcia \emph{et al.}, 1990).

SSVEP measures of thresholds for detecting purely chromatic stimuli find
responses to isoluminant stimuli by around 5-6 weeks postnatally, with
contrast detection reaching near-adult levels by around six months
(Morrone \emph{et al.}, 1993). Spatial acuity as measured by SSVEP
reaches adult levels more slowly, but near-adult levels are recorded
around one year (Norcia \& Tyler, 1985\emph{b}; Hamer \emph{et al.},
1989) compared to around six to seven years with behavioural measures
(Atkinson \& Braddick, 1983; Ellemberg \emph{et al.}, 1999).

At least some of this difference is likely due to the relative
objectivity and high SNR of the SSVEP technique compared to other
methods such as preferential looking, which requires careful measurement
of the infant's gaze direction. In slightly older chilren, other groups
have reported electrophysiological correlates of visual acuity that more
closely match the behavioural measures (De Vries-Khoe \& Spekreijse,
1982).

The SSVEP technique has also been used to study the development of the
contrast gain control mechanisms described in the previous section.
Although response gain (Morrone \& Burr, 1986) or contrast gain control
(Skoczenski \& Norcia, 1998; Candy \emph{et al.}, 2001) is measurable in
infants as young as six weeks old, its development appears to be slower,
with adult levels being reached at approximately 11 years (Pei \emph{et
al.}, 2017).

@@

Amblyopia is a disorder of binocular vision, characterised by one eye
contributing much less to perception than the other. This is often due
to strabismus (squint) or anisometropia (difference in optical
prescription between the eyes) during development. Contemporary accounts
suggest that the amblyopic eye is suppressed by signals from the fellow
eye. SSVEPs provide a convenient and objective method to characterise
the difference in neural response to signals in each eye, and typically
show reduced responses to stimuli in the amblyopic eye (see
Figure~\ref{fig-plotclinical}b) across the contrast range (Baker
\emph{et al.}, 2015; Lygo \emph{et al.}, 2021). Measurements of
interocular contrast modulation (for example, Norcia \emph{et al.},
2000; Hou 侯川 \emph{et al.}, 2021) can also provide information about
the way that amblyopia changes the balance of excitation and inhibition
between eyes. There are currently many novel binocular treatments for
amblyopia under development, often involving virtual reality or stereo
display systems designed to encourage the two eyes to work together. The
steady-state approach may be more sensitive and objective than typical
acuity measurements, and also has the potential to measure suppression
between the eyes directly (e.g. Zheng \emph{et al.}, 2019; Hu \emph{et
al.}, 2023; Du \emph{et al.}, 2023).

@@

\includegraphics{Figures/7_clinicalplot.pdf}

@@

\bibitem[\citeproctext]{ref-bobakCorticalContrastGain1988a}
Bobak P, Bodis-Wollner I \& Marx MS (1988).
\href{https://doi.org/10.1113/jphysiol.1988.sp017340}{Cortical contrast
gain control in human spatial vision.} \emph{The Journal of Physiology}
\textbf{405,} 421--437.

@@

\bibitem[\citeproctext]{ref-ohzawaContrastGainControl1982}
Ohzawa I, Sclar G \& Freeman RD (1982).
\href{https://doi.org/10.1038/298266a0}{Contrast gain control in the cat
visual cortex}. \emph{Nature} \textbf{298,} 266--268.

\bibitem[\citeproctext]{ref-ohzawaContrastGainControl1985}
Ohzawa I, Sclar G \& Freeman RD (1985).
\href{https://doi.org/10.1152/jn.1985.54.3.651}{Contrast gain control in
the cat's visual system}. \emph{Journal of Neurophysiology} \textbf{54,}
651--667.

@@

\bibitem[\citeproctext]{ref-reganRapidObjectiveRefraction1973}
Regan D (1973\emph{b}). Rapid {Objective Refraction Using Evoked Brain
Potentials}. \emph{Investigative Ophthalmology \& Visual Science}
\textbf{12,} 669--679.
